# Models With Memory

Load the packages.

```{r}
#| message: false
#| warning: false

# Load
library(tidyverse)
library(tidybayes)
library(rstan)
library(loo)
library(patchwork)
library(bayesplot)
library(posterior)

# Drop grid lines
theme_set(
  theme_gray() +
    theme(panel.grid = element_blank())
)
```

#### Rethinking: A model by any other name.

## Example: Multilevel tadpoles {#sec-Example-Multilevel-tadpoles}

Let's load the `reedfrogs` data [see @voneshCompensatoryLarvalResponses2005].

```{r}
data(reedfrogs, package = "rethinking")
d <- reedfrogs |>
  # Add a the `tank` cluster variable
  mutate(tank = 1:n())
rm(reedfrogs)

# What?
glimpse(d)
```

Here's the formula for the un-pooled model in which each `tank` gets its own intercept:

$$
\begin{align*}
\text{surv}_i             & \sim \operatorname{Binomial}(n_i, p_i) \\
\operatorname{logit}(p_i) & = \alpha_{\text{tank}[i]} \\
\alpha_j                  & \sim \operatorname{Normal} (0, 1.5) & \text{for } j = 1, \dots, 48,
\end{align*}
$$

where $n_i$ is indexed by the `density` column. Its values are distributed like so.

```{r}
d |> 
  count(density)
```

Now we make the `stan_data` with the `compose_data()` function.

```{r}
stan_data <- d |>
  select(surv, density, tank) |> 
  compose_data()

# What?
str(stan_data)
```

Make the `model_code` for the first simple model, which follows the basic format we learned back in @sec-Aggregated-binomial-Chimpanzees-again-condensed. Also note how we're getting ready for information criteria with the `generated quantities` block.

```{r}
model_code_13.1 <- '
data {
  int<lower=1> n;                       
  array[n] int tank;
  array[n] int density;
  array[n] int surv;
}
parameters {
  vector[n] a;
}
model {
  surv ~ binomial(density, inv_logit(a[tank]));
  a ~ normal(0, 1.5);
}
generated quantities {
  vector[n] log_lik;
  for (i in 1:n) log_lik[i] = binomial_lpmf(surv[i] | density[i], inv_logit(a[tank[i]]));
}
'
```

Now fit the model with `stan()`.

```{r}
#| echo: false

# save(m13.1, file = "fits/m13.1.rda")
load(file = "fits/m13.1.rda")
```

```{r}
#| eval: false
m13.1 <- stan(
  model_code = model_code_13.1, 
  data = stan_data,
  cores = 4, seed = 13)
```

Check the model summary.

```{r}
print(m13.1, pars = c("a[1]", "a[48]"), probs = c(0.055, 0.945))
```

Note how I restricted the output to the first and last intercept with the `pars` argument. You can adjust that code to inspect more or fewer of the intercepts, as desired.

Now consider the multilevel alternative. Its formula is

$$
\begin{align*}
\text{surv}_i             & \sim \operatorname{Binomial}(n_i, p_i) \\
\operatorname{logit}(p_i) & = \alpha_{\text{tank}[i]} \\
\alpha_j                  & \sim \operatorname{Normal}(\color{blue}{\bar \alpha}, \color{blue} \sigma) \\
\color{blue}{\bar \alpha} & \sim \color{blue}{\operatorname{Normal}(0, 1.5)} \\
\color{blue} \sigma       & \sim \color{blue}{\operatorname{Exponential}(1)},
\end{align*}
$$

where

> the prior for the tank intercepts is now a function of two parameters, $\bar \alpha$ and $\sigma$. You can say $\bar \alpha$ like "bar alpha." The bar means average. These two parameters inside the prior is where the "multi" in multilevel arises. The Gaussian distribution with mean $\bar \alpha$ standard deviation $\sigma$ is the prior for each tank’s intercept. But that prior itself has priors for $\bar \alpha$ and $\sigma$. So there are two levels in the model, each resembling a simpler model. (p. 403, *emphasis* in the original)

Here we make the `model_code_13.2`, and fit the model with `stan()`.

```{r}
#| echo: false

# save(m13.2, file = "fits/m13.2.rda")
load(file = "fits/m13.2.rda")
```

```{r}
#| eval: false

model_code_13.2 <- '
data {
  int<lower=1> n;                       
  array[n] int tank;
  array[n] int density;
  array[n] int surv;
}
parameters {
  real abar;
  vector[n] a;
  real<lower=0> sigma;
}
model {
  surv ~ binomial(density, inv_logit(a[tank]));
  a ~ normal(abar, sigma);
  abar ~ normal(0, 1.5);
  sigma ~ exponential(1);
}
generated quantities {
  vector[n] log_lik;
  for (i in 1:n) log_lik[i] = binomial_lpmf(surv[i] | density[i], inv_logit(a[tank[i]]));
}
'

m13.2 <- stan(
  model_code = model_code_13.2, 
  data = stan_data,
  cores = 4, seed = 13)
```

Check the model summary.

```{r}
print(m13.2, pars = c("abar", "sigma", "a[1]", "a[48]"), probs = c(0.055, 0.945))
```

Compare the two models by the WAIC with the `extract_log_lik()`, `waic()`, and `loo_compare()` functions.

```{r}
#| warning: false

loo_compare(
  extract_log_lik(m13.1) |> waic(),
  extract_log_lik(m13.2) |> waic()
) |> 
  print(simplify = FALSE)
```

Our results match up nicely with those in the text (p. 404). Here are the stacking weights.

```{r}
#| warning: false

set.seed(13)

loo_model_weights(
  list(extract_log_lik(m13.1) |> loo(),
       extract_log_lik(m13.2) |> loo()),
  method = "stacking"
)
```

All the weight goes to the multilevel model.

Here we use a little `bind_rows()` and `spread_draws()` magic to make a version of Figure 13.1 based on both models.

```{r}
#| fig-width: 6
#| fig-height: 5.25

bind_rows(
  m13.1 |> spread_draws(a[tank]),
  m13.2 |> spread_draws(a[tank])
) |> 
  mutate(model = rep(c("conventional (m13.1)", "multilevel (m13.2)"), each = n() / 2),
         propsurv = plogis(a)) |> 
  group_by(tank, model) |> 
  mean_qi(propsurv, .width = 0.89) |> 
  left_join(d |> 
              select(tank, density), 
            by = join_by(tank)) |> 
  
  ggplot(aes(x = tank, y = propsurv)) +
  geom_hline(yintercept = mean(d$propsurv), color = "white") +
  geom_pointinterval(aes(ymin = .lower, ymax = .upper),
                     linewidth = 1/4, shape = 1) +
  geom_point(data = d,
             color = "blue") +
  scale_y_continuous("proportion survival", limits = 0:1) +
  facet_grid(model ~ density, labeller = labeller(density = label_both), scales = "free_x")
```

Here's a version of Figure 13.2.

```{r}
#| fig-width: 6
#| fig-height: 3
#| warning: false

set.seed(13)

# Left
p1 <- as_draws_df(m13.2) |> 
  select(.draw, abar, sigma) |> 
  slice_sample(n = 100) |> 
  expand_grid(x = seq(from = -3.5, to = 4.5, length.out = 101)) |> 
  mutate(density = dnorm(x = x, mean = abar, sd = sigma)) |> 
  
  ggplot(aes(x = x, y = density, group = .draw)) +
  geom_line(alpha = 1/3, linewidth = 1/3) +
  xlab("log-odds survive") +
  coord_cartesian(xlim = c(-3, 4))

# Right
p2 <- as_draws_df(m13.2) |> 
  select(.draw, abar, sigma) |> 
  slice_sample(n = 8000, replace = TRUE) |> 
  mutate(sim_tanks = rnorm(n = n(), mean = abar, sd = sigma)) |> 
  mutate(p = plogis(sim_tanks)) |> 
  
  ggplot(aes(x = p)) +
  geom_density(adjust = 0.1, fill = "gray65", linewidth = 0) +
  labs(x = "probability survive",
       y = NULL)

# Combine
p1 | p2
```

#### Overthinking: QUAP fails, MCMC succeeds.

#### Rethinking: Varying intercepts as over-dispersion.

#### Overthinking: Prior for variance components.

Here's how we define and fit the model with the half-normal prior on $\sigma$.

```{r}
#| echo: false

# save(m13.2b, file = "fits/m13.2b.rda")
load(file = "fits/m13.2b.rda")
```

```{r}
#| eval: false

model_code_13.2b <- '
data {
  int<lower=1> n;                       
  array[n] int tank;
  array[n] int density;
  array[n] int surv;
}
parameters {
  real abar;
  vector[n] a;
  real<lower=0> sigma;
}
model {
  surv ~ binomial(density, inv_logit(a[tank]));
  a ~ normal(abar, sigma);
  abar ~ normal(0, 1.5);
  sigma ~ normal(0, 1);
}
'

m13.2b <- stan(
  model_code = model_code_13.2b, 
  data = stan_data,
  cores = 4, seed = 13)
```

Check the model summary.

```{r}
print(m13.2b, pars = c("abar", "sigma", "a[1]", "a[48]"), probs = c(0.055, 0.945))
```

If you're curious how the exponential and half-Normal priors compare to one another and to their posteriors, you might just plot.

```{r}
#| fig-width: 5.5
#| fig-height: 2.5

# For descriptive y-axis labels
prior_vec <- c("sigma%~%exponential(1)", "sigma%~%'half-normal(0, 1)'")

# Wrangle
bind_rows(as_draws_df(m13.2), as_draws_df(m13.2b)) |> 
  mutate(prior = rep(prior_vec, each = n() / 2)) |> 
  
  # Plot
  ggplot(aes(x = sigma, y = prior)) +
  stat_halfeye(point_interval = mean_qi, .width = 0.89,
               linewidth = 1, shape = 1) +
  scale_y_discrete(NULL, expand = expansion(mult = 0.1), 
                   labels = ggplot2:::parse_safe) +
  xlab(expression(sigma[posterior]))
```

In this case, the prior didn't matter much.

## Varying effects and the underfitting/overfitting trade-off

### The model.

The simulation formula should look familiar.

$$
\begin{align*}
\text{surv}_i & \sim \operatorname{Binomial}(n_i, p_i) \\
\operatorname{logit}(p_i) & = \alpha_{\text{pond}[i]} \\
\alpha_j                  & \sim \operatorname{Normal}(\bar \alpha, \sigma) \\
\bar \alpha               & \sim \operatorname{Normal}(0, 1.5) \\
\sigma                    & \sim \operatorname{Exponential}(1)
\end{align*}
$$

### Assign values to the parameters.

Here we follow along with McElreath and "assign specific values representative of the actual tadpole data" (p. 409). Because he included a `set.seed()` line in his **R** code 13.8, our results should match his exactly.

```{r}
a_bar   <-  1.5
sigma   <-  1.5
n_ponds <- 60

set.seed(5005)

dsim <- tibble(
  pond   = 1:n_ponds,
  ni     = rep(c(5, 10, 25, 35), each = n_ponds / 4) |> as.integer(),
  true_a = rnorm(n = n_ponds, mean = a_bar, sd = sigma))

# What?
glimpse(dsim)
```

McElreath twice urged us to inspect the contents of this simulation. In addition to looking at the data with `glimpse()`, we might well plot.

```{r}
#| fig-width: 5
#| fig-height: 3

dsim |> 
  mutate(ni = factor(ni)) |> 
  
  ggplot(aes(x = true_a, y = ni)) +
  stat_dotsinterval(size = 2/3, slab_size = 0, .width = 0.5) +
  ggtitle("Log-odds varying by # tadpoles per pond") +
  theme(plot.title = element_text(size = 14))
```

#### Overthinking: Data types and Stan models.

> There are two basic types of numerical data in R, in- tegers and real values. A number like "3" could be either. Inside your computer, integers and real ("numeric") values are represented differently. For example, here is the same vector of values generated as both:

```{r}
class(1:3)
class(c(1, 2, 3))
```

> Usually, you don't have to manage these types, because R manages them for you. But when you pass values to Stan, or another external program, often the internal representation does matter. In particular, Stan and ~~`ulam`~~ [`stan()`] sometimes require explicit integers. For example, in a binomial model, the "size" variable that specifies the number of trials must be of integer type. Stan may provide a mysterious warning message about a function not being found, when the size variable is instead of "real" type, or what R calls `numeric`. Using `as.integer` before passing the data to Stan or ~~`ulam`~~ [`stan()`] will resolve the issue. (p. 410)

### Sumulate survivors.

> Each pond $i$ has $n_i$ potential survivors, and nature flips each tadpole's coin, so to speak, with probability of survival $p_i$. This probability $p_i$ is implied by the model definition, and is equal to:
>
> $$p_i = \frac{\exp (\alpha_i)}{1 + \exp (\alpha_i)}$$
>
> The model uses a logit link, and so the probability is defined by the [`plogis()`] function. (p. 411)

Although McElreath shared his `set.seed()` number in the last section, he didn't share it for this bit. We'll go ahead and carry over the one from last time. However, in a moment we'll see this clearly wasn't the one he used here. As a consequence, our results will deviate a bit from his.

```{r}
set.seed(5005)

dsim <- dsim |>
  mutate(si = rbinom(n = n(), prob = plogis(true_a), size = ni))

# What?
glimpse(dsim)
```

### Compute the no-pooling estimates.

The no-pooling estimates (i.e., $\alpha_{\text{tank}[i]}$) are the results of simple algebra.

```{r}
dsim <- dsim |>
  mutate(p_nopool = si / ni)

# What?
glimpse(dsim)
```

"These are the same no-pooling estimates you'd get by fitting a model with a dummy variable for each pond and flat priors that induce no regularization" (p. 411). That is, these are the same kinds of estimates we got back when we fit `m13.1`.

### Compute the partial-pooling estimates.

Now we make the `stan_data` with the `compose_data()` function.

```{r}
stan_data <- dsim |>
  select(si, ni, pond) |> 
  compose_data()

# What?
str(stan_data)
```

Here we define and fit the model with `stan()`.

```{r}
#| echo: false

# save(m13.3, file = "fits/m13.3.rda")
load(file = "fits/m13.3.rda")
```

```{r}
#| eval: false

model_code_13.3 <- '
data {
  int<lower=1> n;                       
  array[n] int pond;
  array[n] int ni;
  array[n] int si;
}
parameters {
  real a_bar;
  vector[n] a_pond;
  real<lower=0> sigma;
}
model {
  si ~ binomial(ni, inv_logit(a_pond[pond]));
  a_pond ~ normal(a_bar, sigma);
  a_bar ~ normal(0, 1.5);
  sigma ~ exponential(1);
}
'

m13.3 <- stan(
  model_code = model_code_13.3, 
  data = stan_data,
  cores = 4, seed = 13)
```

Check the model summary.

```{r}
print(m13.3, 
      pars = c("a_pond[1]", "a_pond[2]", 
               "a_pond[59]", "a_pond[60]",
               "a_bar", "sigma"), 
      probs = c(0.055, 0.945))
```

To make Figure 13.3, we'll first need a couple data objects. The `draws` data frame has the information for the points, and the `draws_means` data frame has the information for the horizontal lines.

```{r}
#| message: false

draws <- m13.3 |> 
  spread_draws(a_pond[pond]) |> 
  mutate(p_partpool = plogis(a_pond)) |> 
  group_by(pond) |> 
  summarise(p_partpool = mean(p_partpool)) |> 
  left_join(dsim |> 
              select(pond, true_a, p_nopool, ni)) |> 
  mutate(p_true = plogis(true_a)) |>
  mutate(nopool_error   = abs(p_nopool   - p_true),
         partpool_error = abs(p_partpool - p_true)) |> 
  pivot_longer(contains("error"), values_to = "error") |> 
  mutate(pooling = ifelse(name == "nopool_error", "no", "partial"))

draws_means <- draws |> 
  group_by(pooling, ni) |> 
  summarise(mean = mean(error))

# What?
glimpse(draws)
glimpse(draws_means)
```

Now make the figure.

```{r}
#| fig-width: 7
#| fig-height: 3.5

draws |> 
  ggplot(aes(x = pond, y = error)) +
  geom_point(aes(color = pooling, shape = pooling)) +
  geom_hline(data = draws_means,
             aes(yintercept = mean, color = pooling, linetype = pooling),
             linewidth = 1/3) +
  scale_color_viridis_d("Pooling?", option = "A", end = 0.6, direction = -1) +
  scale_linetype_manual("Pooling?", values = 1:2) +
  scale_shape_manual("Pooling?", values = c(19, 1)) +
  ylab("absolute error") +
  facet_wrap(~ ni, labeller = label_both, nrow = 1, scales = "free_x")
```

#### Overthinking: Repeating the pond simulation.

Make the new `new_dsim` data frame, and them update the `stan_data`.

```{r}
set.seed(1999)  # For new data, set a new seed

new_dsim <- tibble(
  pond   = 1:n_ponds,
  ni     = rep(c(5, 10, 25, 35), each = n_ponds / 4) |> as.integer(),
  true_a = rnorm(n = n_ponds, mean = a_bar, sd = sigma)) |> 
  mutate(si = rbinom(n = n(), prob = plogis(true_a), size = ni),
         p_nopool = si / ni)

stan_data <- new_dsim |>
  select(si, ni, pond) |> 
  compose_data()
```

Now update the fit with the `sampling()` function, saving the results as `m13.3new`.

```{r}
#| echo: false

# save(m13.3new, file = "fits/m13.3new.rda")
load(file = "fits/m13.3new.rda")
```

```{r}
#| eval: false

m13.3new <- sampling(
    object = m13.3@stanmodel,
    data = stan_data,
    cores = 4, seed = 13) 
```

Make the new variant of the figure.

```{r}
#| fig-width: 7
#| fig-height: 3.5
#| message: false

# Wrangle
# Use `m13.3new` in place of `m13.3`
draws_new <- m13.3new |> 
  spread_draws(a_pond[pond]) |> 
  mutate(p_partpool = plogis(a_pond)) |> 
  group_by(pond) |> 
  summarise(p_partpool = mean(p_partpool)) |> 
  # Use `new_dsim` in place of `new`
  left_join(new_dsim |> 
              select(pond, true_a, p_nopool, ni)) |> 
  mutate(p_true = plogis(true_a)) |>
  mutate(nopool_error   = abs(p_nopool   - p_true),
         partpool_error = abs(p_partpool - p_true)) |> 
  pivot_longer(contains("error"), values_to = "error") |> 
  mutate(pooling = ifelse(name == "nopool_error", "no", "partial"))

draws_means_new <- draws_new |> 
  group_by(pooling, ni) |> 
  summarise(mean = mean(error))

# Plot
draws_new |> 
  ggplot(aes(x = pond, y = error)) +
  geom_point(aes(color = pooling, shape = pooling)) +
  geom_hline(data = draws_means_new,
             aes(yintercept = mean, color = pooling, linetype = pooling),
             linewidth = 1/3) +
  scale_color_viridis_d("Pooling?", option = "D", end = 0.6, direction = -1) +
  scale_linetype_manual("Pooling?", values = 1:2) +
  scale_shape_manual("Pooling?", values = c(19, 1)) +
  ylab("absolute error") +
  facet_wrap(~ ni, labeller = label_both, nrow = 1, scales = "free_x")
```

## More than one type of cluster {#sec-More-than-one-type-of-cluster}

#### Rethinking: Cross-classification and hierarchy.

### Multilevel chimpanzees.

The initial multilevel update from model `m11.4` from @sec-Logistic-regression-Prosocial-chimpanzees follows the statistical formula

```{r}
#| eval: false
#| echo: false

pulled_left
pulled-left
```

$$
\begin{align*}
\text{pulled-left}_i         & \sim \operatorname{Binomial}(n_i = 1, p_i) \\
\operatorname{logit} (p_i) & = \alpha_{\text{actor}[i]} + \beta_{\text{treatment}[i]} + \color{blue}{\gamma_{\text{block}[i]}} \\
\alpha_j & \sim \operatorname{Normal}(\bar \alpha, \sigma_\alpha), \;\;\; \text{for } j = 1, \dots, 7 \\
\beta_j  & \sim \operatorname{Normal}(0, 0.5), \;\;\; \text{for } j = 1, \dots, 4 \\
\color{blue}{\gamma_j} & \sim \color{blue}{\operatorname{Normal}(0, \sigma_\gamma), \;\;\; \text{for } j = 1, \dots, 6} \\
\bar \alpha   & \sim \operatorname{Normal}(0, 1.5) \\
\sigma_\alpha & \sim \operatorname{Exponential}(1) \\
\color{blue}{\sigma_\gamma} & \sim \color{blue}{\operatorname{Exponential}(1)}.
\end{align*}
$$

Load the `chimpanzees` data, wrangle, and make the `stan_data`.

```{r}
data(chimpanzees, package = "rethinking")

d <- chimpanzees |> 
  mutate(actor     = factor(actor),
         block     = factor(block),
         treatment = factor(1 + prosoc_left + 2 * condition))

rm(chimpanzees)

stan_data <- d |>
  select(pulled_left, actor, block, treatment) |>
  compose_data()

# What?
str(stan_data)
```

Define `model_code_13.4`.

```{r}
model_code_13.4 <- '
data {
  int<lower=1> n;                       
  int<lower=1> n_actor;
  int<lower=1> n_treatment;
  int<lower=1> n_block;
  array[n] int actor;
  array[n] int treatment;
  array[n] int block;
  array[n] int<lower=0, upper=1> pulled_left;
}
parameters {
  vector[n_actor] a;
  vector[n_treatment] b;
  vector[n_block] g;
  
  real a_bar;
  real<lower=0> sigma_a;
  real<lower=0> sigma_g;
}
model {
  pulled_left ~ binomial(1, inv_logit(a[actor] + b[treatment] + g[block]));
  
  b ~ normal(0, 0.5);
  // adaptive priors
  a ~ normal(a_bar, sigma_a);
  g ~ normal(0, sigma_g);
  // hyper-priors
  a_bar ~ normal(0, 1.5);
  [sigma_a, sigma_g] ~ exponential(1);
}
generated quantities {
  vector[n] log_lik;
  for (i in 1:n) log_lik[i] = binomial_lpmf(pulled_left[i] | 1, inv_logit(a[actor[i]] + b[treatment[i]] + g[block[i]]));
}
'
```

Fit the cross-classified chimp model with `stan()`.

```{r}
#| echo: false

# save(m13.4, file = "fits/m13.4.rda")
load(file = "fits/m13.4.rda")
```

```{r}
#| eval: false

m13.4 <- stan(
  model_code = model_code_13.4, 
  data = stan_data,
  cores = 4, seed = 13)
```

Like in the text, this model returned warnings about divergent transitions, which isn't great. As we learned back in @sec-Non-identifiable-parameters, we can check for the divergent-transition warnings with the `check_divergences()` function.

```{r}
check_divergences(m13.4)
```

Here's the parameter summary.

```{r}
print(m13.4, 
      include = FALSE,
      pars = c("log_lik", "lp__"),
      probs = c(0.055, 0.945))
```

Here's Figure 13.4.

```{r}
#| fig-width: 7
#| fig-height: 4.5
#| warning: false

# For the y-axis ordering
parameter_vec <- c(str_c("b[", 1:4, "]"), str_c("a[", 1:7, "]"), str_c("g[", 1:6, "]"), "a_bar", str_c("sigma_", c("a", "g")))

# Left panel
p1 <- as_draws_df(m13.4) |>
  select(-contains("log_lik")) |> 
  pivot_longer(cols = `a[1]`:sigma_g) |> 
  mutate(name = factor(name, levels = parameter_vec) |> 
           fct_rev()) |> 
  
  ggplot(aes(x = value, y = name)) +
  geom_vline(xintercept = 0, color = "white") +
  stat_pointinterval(.width = 0.89, linewidth = 1, shape = 1) +
  ylab(NULL)

# For annotation
d_text <- tibble(
  group = c("actor", "block"),
  value = c(3, 0.6),
  y = c(0.10, 0.75))

# Right panel
p2 <- as_draws_df(m13.4) |> 
  select(sigma_a, sigma_g) |> 
  pivot_longer(everything()) |> 
  mutate(group = ifelse(name == "sigma_a", "actor", "block")) |> 
  
  ggplot(aes(x = value)) +
  stat_histinterval(aes(fill = group, group = group),
                    .width = 0.89, alpha = 0.9, linewidth = 1, shape = 1) +
  geom_text(data = d_text,
            aes(y = y, label = group, color = group)) +
  scale_y_continuous(NULL, breaks = NULL) +
  scale_color_viridis_d(option = "C", begin = 0.4, end = 0.7, breaks = NULL) +
  scale_fill_viridis_d(option = "C", begin = 0.4, end = 0.7, breaks = NULL) +
  xlab("standard deviation") +
  coord_cartesian(xlim = c(0, 4))

# Define the `layout` for the panels
layout <- c(
  area(t = 1, b = 4, l = 1, r = 1),
  area(t = 2, b = 4, l = 2, r = 2))

# Combine and display
p1 + p2 + 
  plot_layout(design = layout)
```

Now define and fit the simpler version of the model that omits the $\gamma$-related parameters, `m13.5`.

```{r}
#| echo: false

# save(m13.5, file = "fits/m13.5.rda")
load(file = "fits/m13.5.rda")
```

```{r}
#| eval: false

model_code_13.5 <- '
data {
  int<lower=1> n;                       
  int<lower=1> n_actor;
  int<lower=1> n_treatment;
  array[n] int actor;
  array[n] int treatment;
  array[n] int<lower=0, upper=1> pulled_left;
}
parameters {
  vector[n_actor] a;
  vector[n_treatment] b;
  real a_bar;
  real<lower=0> sigma_a;
}
model {
  pulled_left ~ binomial(1, inv_logit(a[actor] + b[treatment]));
  b ~ normal(0, 0.5);
  // adaptive prior
  a ~ normal(a_bar, sigma_a);
  // hyper-priors
  a_bar ~ normal(0, 1.5);
  sigma_a ~ exponential(1);
}
generated quantities {
  vector[n] log_lik;
  for (i in 1:n) log_lik[i] = binomial_lpmf(pulled_left[i] | 1, inv_logit(a[actor[i]] + b[treatment[i]]));
}
'

m13.5 <- stan(
  model_code = model_code_13.5, 
  data = stan_data,
  cores = 4, seed = 13)
```

Compare the two models by the WAIC.

```{r}
#| warning: false

loo_compare(
  extract_log_lik(m13.4) |> waic(),
  extract_log_lik(m13.5) |> waic()
) |> 
  print(simplify = FALSE)
```

Here are the stacking weights.

```{r}
#| warning: false

set.seed(13)

loo_model_weights(
  list(extract_log_lik(m13.4) |> loo(),
       extract_log_lik(m13.5) |> loo()),
  method = "stacking"
)
```

The stacking weights more heavily favor the simpler model `m13.5`, than the WAIC weights McElreath reported in the book (p. 418).

> There is nothing to gain here by selecting either model. The comparison of the two models tells a richer story... Since this is an experiment, there is nothing to really select. The experimental design tells us the relevant causal model to inspect. (pp. 418--419)

### Even more clusters.

The next model `m13.6` is a minor generalization of `m13.4`, where the $\beta$ parameters are also partially pooled.

```{r}
#| echo: false

# save(m13.6, file = "fits/m13.6.rda")
load(file = "fits/m13.6.rda")
```

```{r}
#| eval: false

model_code_13.6 <- '
data {
  int<lower=1> n;                       
  int<lower=1> n_actor;
  int<lower=1> n_treatment;
  int<lower=1> n_block;
  array[n] int actor;
  array[n] int treatment;
  array[n] int block;
  array[n] int<lower=0, upper=1> pulled_left;
}
parameters {
  vector[n_actor] a;
  vector[n_treatment] b;
  vector[n_block] g;
  real a_bar;
  real<lower=0> sigma_a;
  real<lower=0> sigma_b;
  real<lower=0> sigma_g;
}
model {
  pulled_left ~ binomial(1, inv_logit(a[actor] + b[treatment] + g[block]));
  // adaptive priors
  a ~ normal(a_bar, sigma_a);
  b ~ normal(0, sigma_b);
  g ~ normal(0, sigma_g);
  // hyper-priors
  a_bar ~ normal(0, 1.5);
  [sigma_a, sigma_b, sigma_g] ~ exponential(1);
}
generated quantities {
  vector[n] log_lik;
  for (i in 1:n) log_lik[i] = binomial_lpmf(pulled_left[i] | 1, inv_logit(a[actor[i]] + b[treatment[i]] + g[block[i]]));
}
'

m13.6 <- stan(
  model_code = model_code_13.6, 
  data = stan_data,
  cores = 4, seed = 13)
```

In place of McElreath's `coeftab()` approach, let's just make a coefficient plot for the $\beta$ parameters.

```{r}
#| fig-width: 7
#| fig-height: 1.75

bind_rows(spread_draws(m13.4, b[j]), 
          spread_draws(m13.6, b[j])) |> 
  mutate(model = rep(c("m13.4", "m13.6"), each = n() / 2),
         parameter = str_c("beta[", j, "]")) |> 
  
  ggplot(aes(x = b, y = parameter, color = model)) +
  stat_pointinterval(point_interval = mean_qi, .width = 0.89, 
                     linewidth = 1, shape = 1,
                     position = position_dodge(width = -0.4)) +
  scale_y_discrete(NULL, labels = ggplot2:::parse_safe) +
  scale_color_viridis_d(NULL, option = "D", end = 0.4) +
  xlab("posterior")
```

As McElreath wrote:

> These are not identical, but they are very close. If you look at `sigma_b`, you'll see that it is small. (p. 420)

Here's `sigma_b`.

```{r}
#| fig-width: 4.5
#| fig-height: 2.5

as_draws_df(m13.6) |> 
  ggplot(aes(x = sigma_b)) +
  stat_halfeye(.width = 0.89, adjust = 1/2, linewidth = 1, shape = 1) +
  scale_y_continuous(NULL, breaks = NULL) +
  xlab(expression(sigma[beta]))
```

That's an impressively jagged posterior.

Compare the two models `m13.4` and `m13.6` by the LOO and you'll see they're almost the same.

```{r}
#| warning: false

loo_compare(
  extract_log_lik(m13.4) |> loo(),
  extract_log_lik(m13.6) |> loo()
) |> 
  print(simplify = FALSE)
```

If you check the `check_divergences()` output, you'll see we still have several divergent transitions with `m13.6`.

```{r}
check_divergences(m13.6)
```

## Divergent transitions and non-centered priors {#sec-Divergent-transitions-and-non-centered-priors}

### Rethinking: No free samples.

### The Devil's Funnel.

McElreath posed a joint distribution

$$
\begin{align*}
v & \sim \operatorname{Normal}(0, 3) \\
x & \sim \operatorname{Normal}(0, \exp(v)),
\end{align*}
$$

where the scale of $x$ depends on another variable, $v$. Here's how we might sample from those priors with `stan()`.

```{r}
#| echo: false

# save(m13.7, file = "fits/m13.7.rda")
load(file = "fits/m13.7.rda")
```

```{r}
#| eval: false

model_code_13.7 <- '
parameters {
  real v;
  real x;
}
model {
  v ~ normal(0, 3);
  x ~ normal(0, exp(v));
}
'

m13.7 <- stan(
  model_code = model_code_13.7, 
  cores = 4, seed = 13)
```

Behold the `print()` summary.

```{r}
print(m13.7, probs = c(0.055, 0.945))
```

We have lots of divergent transitions, and other scary warning messages, too.

```{r}
check_hmc_diagnostics(m13.7)
```

Here's a scatter plot of the results.

```{r}
#| fig-width: 3
#| fig-height: 2.75

as_draws_df(m13.7) |> 
  ggplot(aes(x = x, y = v)) +
  geom_point(alpha = 1/2, size = 1/4) +
  labs(x = expression(italic(x)),
       y = expression(italic(v)))
```

This is something of a sample version of McElreath's Figure 13.5a.

Here are the trace plots via `mcmc_trace()`. They're a mess.

```{r}
#| fig-width: 6
#| fig-height: 3

mcmc_trace(m13.7, 
           pars = vars(v:x),
           facet_args = list(ncol = 1))
```

To avoid the divergent transitions than can arise models using the centered parameterization, we can switch from our original formula to a non-centered parameterization, such as:

$$
\begin{align*}
v & \sim \operatorname{Normal}(0, 3) \\
z & \sim \operatorname{Normal}(0, 1) \\
x & = z \exp(v),
\end{align*}
$$

where $x$ is now a deterministic function of two independent distributions, $v$ and $z$. For our `stan()`-based workflow, this means defining `v` and `z` in the `parameters` block, and then defining `x` as a function of the two in the `generated quantities`. But because `x` is a deterministic function of `v` and `z`, only `v` and `z` get assigned priors in the `model` block.

```{r}
#| echo: false

# save(m13.7nc, file = "fits/m13.7nc.rda")
load(file = "fits/m13.7nc.rda")
```

```{r}
#| eval: false

model_code_13.7nc <- '
parameters {
  real v;
  real z;
}
model {
  v ~ normal(0, 3);
  z ~ normal(0, 1);
}
generated quantities {
  real x = z * exp(v);
}
'

m13.7nc <- stan(
  model_code = model_code_13.7nc, 
  cores = 4, seed = 13)
```

Behold the `print()` summary.

```{r}
print(m13.7nc, probs = c(0.055, 0.945))
```

The values in the `n_eff` and `Rhat` columns are much better.

Here's a scatter plot of the new results, which is something of a version of Figure 13.5b.

```{r}
#| fig-width: 3
#| fig-height: 2.75

as_draws_df(m13.7nc) |> 
  ggplot(aes(x = z, y = v)) +
  geom_point(alpha = 1/2, size = 1/4) +
  labs(x = expression(italic(z)),
       y = expression(italic(v)))
```

The trace plots for $v$ and $z$ look pretty great, too.

```{r}
#| fig-width: 6
#| fig-height: 4.5

mcmc_trace(m13.7nc, 
           pars = vars(v, z, x),
           facet_args = list(ncol = 1))
```

The trace plot for $x$, however, is a mess. This isn't necessarily a problem, because we're only really interested in $v$ and $z$. $x$ is a means to an end.

### Non-centered chimpanzees. {#sec-Non-centered-chimpanzees}

At the top of the section, McElreath reported the `rethinking::ulam()` default is to set `adapt_delta = 0.95`. Readers should be aware that the `rstan::stan()` default is `adapt_delta = 0.80`. See the `stan` section of the *rstan reference manual* (link [here](https://CRAN.R-project.org/package=rstan/rstan.pdf)). A consequence of this difference is `rethinking::ulam()` will tend to take smaller step sizes than `rstan::stan()`, at the cost of slower exploration of the posterior. I don't know that one is inherently better than the other. They're just defaults.

We can just use the `sampling()` function to revisit `m13.4`. This time, we increase the `adapt_delta` parameter in a list fed to the `control` argument.

```{r}
#| warning: false

m13.4b <- sampling(
  object = m13.4@stanmodel,
  data = stan_data,
  cores = 4, seed = 13,
  control = list(adapt_delta = 0.99))
```

We've been using the `check_divergences()` function to check for divergent-transitions. In addition to the number of divergent transitions, that function also returns recommendations for next steps, such as increasing the `adapt_delta` parameter. If all you want is a more focused output with just the number of divergent transitions, you might use the `get_num_divergent()` function instead. Here we compare the `get_num_divergent()` output from both `m13.4` and `m13.4b`.

```{r}
get_num_divergent(m13.4)
get_num_divergent(m13.4b)
```

We dropped from 108 to 4, which isn't nothing.

Here's the `print()` summary.

```{r}
print(m13.4b, 
      include = FALSE,
      pars = c("log_lik", "lp__"),
      probs = c(0.055, 0.945))
```

For `m13.4nc`, we first update the statistical model to

$$
\begin{align*}
\text{pulled-left}_i & \sim \operatorname{Binomial}(n_i = 1, p_i) \\
\operatorname{logit} (p_i) & = \color{blue}{\underbrace{\bar \alpha + z_{\text{actor}[i]} \sigma_\alpha}_{\alpha_{\text{actor}[i]}}} + \beta_{\text{treatment}[i]} \color{blue} + \color{blue}{\underbrace{x_{\text{block}[i]} \sigma_\gamma}_{\gamma_{\text{block}[i]}}} \\
\beta_j  & \sim \operatorname{Normal}(0, 0.5), \;\;\; \text{for } j = 1, \dots, 4 \\
\color{blue}{x_j} & \sim \color{blue}{\operatorname{Normal}(0, 1)} \\
\color{blue}{z_j} & \sim \color{blue}{\operatorname{Normal}(0, 1)} \\
\bar \alpha   & \sim \operatorname{Normal}(0, 1.5) \\
\sigma_\alpha, \sigma_\gamma & \sim \operatorname{Exponential}(1),
\end{align*}
$$

where each actor intercept $\alpha_j$ is defined by $\alpha + z_j \sigma_\alpha$, and each block intercept $\gamma_j$ is defined by $x_j \sigma_\gamma$.

Here's how we might define `model_code_13.4nc`.

```{r}
model_code_13.4nc <- '
data {
  int<lower=1> n;                       
  int<lower=1> n_actor;
  int<lower=1> n_treatment;
  int<lower=1> n_block;
  array[n] int actor;
  array[n] int treatment;
  array[n] int block;
  array[n] int<lower=0, upper=1> pulled_left;
}
parameters {
  vector[n_actor] z;
  vector[n_block] x;
  vector[n_treatment] b;
  real a_bar;
  real<lower=0> sigma_a;
  real<lower=0> sigma_g;
}
model {
  vector[n] eta;  // To simplify the likelihood line
  eta = a_bar + z[actor] * sigma_a + x[block] * sigma_g + b[treatment];
  
  pulled_left ~ binomial(1, inv_logit(eta));
  
  a_bar ~ normal(0, 1.5);
  z ~ normal(0, 1);
  x ~ normal(0, 1);
  [sigma_a, sigma_g] ~ exponential(1);
  b ~ normal(0, 0.5);
}
generated quantities {
  vector[n_actor] a = a_bar + z * sigma_a;
  vector[n_block] g = x * sigma_g;
}
'
```

Now fit the non-centered version of the model with `stan()`.

```{r}
#| echo: false

# save(m13.4nc, file = "fits/m13.4nc.rda")
load(file = "fits/m13.4nc.rda")
```

```{r}
#| eval: false

m13.4nc <- stan(
  model_code = model_code_13.4nc, 
  data = stan_data,
  cores = 4, seed = 13)
```

The `print()` summary for this version of the model is the best one yet.

```{r}
print(m13.4nc, 
      include = FALSE,
      pars = c("log_lik", "lp__"),
      probs = c(0.055, 0.945))
```

We use the `summarise_draws()` function and the `parameter_vec` from a while back to make a version of McElreath's Figure 13.6. Here we replace the old-fashioned `n_eff` values with the newer `ess_bulk` and `ess_tail`.

```{r}
#| fig-height: 3

bind_rows(
  summarise_draws(m13.4, ess_bulk, ess_tail),
  summarise_draws(m13.4nc, ess_bulk, ess_tail)
) |> 
  filter(variable %in% parameter_vec) |> 
  mutate(fit = rep(c("m13.4", "m13.4nc"), each = n() / 2)) |> 
  pivot_longer(cols = contains("ess_")) |> 
  pivot_wider(names_from = fit, values_from = value) |> 
  
  ggplot(aes(x = m13.4, y = m13.4nc)) +
  geom_abline(color = "white") +
  geom_point(shape = 1) +
  coord_equal(xlim = c(0, 5000),
              ylim = c(0, 5000)) +
  labs(x = "m13.4 (centered)",
       y = "m13.4nc (non-centered)") +
  facet_wrap(~ name)
```

Big difference.

## Multilevel posterior predictions

### Posterior prediction for same clusters.

If you are using a by-hand approach for computing model-based expectations for a multilevel model of this kind, `spread_draws()` seems like a good solution for a tidy workflow. Here's what that can look like for `actor == 2`, in `block == 1`. We just display the results in a plot resembling Figure 13.7a.

```{r}
#| fig-width: 2
#| fig-height: 3

chimp <- 2

d_pred <- tibble(
    actor = chimp,
    block = 1,
    treatment = 1:4)

# For the x-axis
treatment_labels_vec <- c("R/N", "L/N", "R/P", "L/P")

m13.4 |> 
  spread_draws(a[actor], b[treatment], g[block]) |> 
  right_join(d_pred, by = join_by(actor, block, treatment)) |> 
  mutate(p = plogis(a + b + g)) |> 
  
  ggplot(aes(x = treatment, y = p)) +
  stat_lineribbon(point_interval = mean_qi, .width = 0.80, 
                  fill = "gray67") +
  scale_x_continuous(labels = treatment_labels_vec) +
  labs(y = "proportion pulled left",
       subtitle = "actor 2")
```

Note the severely-restricted range on the y axis.

Part of the key to the `spread_draws()` approach is we named the indices for the `a`, `b`, and `g` vectors after the corresponding variables in the `d_pred` predictor data, and then used `right_join()` to connect the two.

Here's what this looks like for `actor == 5`, as in Figure 13.7.

```{r}
#| fig-width: 2
#| fig-height: 3

# This has been updated
chimp <- 5

d_pred <- tibble(
    actor = chimp,
    block = 1,
    treatment = 1:4)

m13.4 |> 
  spread_draws(a[actor], b[treatment], g[block]) |> 
  right_join(d_pred, by = join_by(actor, block, treatment)) |> 
  mutate(p = plogis(a + b + g)) |> 
  
  ggplot(aes(x = treatment, y = p)) +
  stat_lineribbon(point_interval = mean_qi, .width = 0.80, 
                  fill = "gray67") +
  scale_x_continuous(labels = treatment_labels_vec) +
  scale_y_continuous("proportion pulled left", limits = 0:1) +
  labs(subtitle = "actor 5")
```

We could, of course, just make a faceted plot of all actors, in all blocks, and all treatments. For such a plot, we don't even need the `d_pred` data.

```{r}
#| fig-width: 8
#| fig-height: 2.75

m13.4 |>
  spread_draws(a[actor], b[treatment], g[block]) |> 
  mutate(p = plogis(a + b + g),
         block = factor(block)) |> 
  
  ggplot(aes(x = treatment, y = p,
             color = block, fill = block, group = block)) +
  stat_lineribbon(point_interval = mean_qi, .width = 0.80, 
                  alpha = 1/3, linewidth = 1) +
  scale_x_continuous(labels = treatment_labels_vec, expand = expansion(mult = 0.1)) +
  scale_y_continuous("proportion pulled left", breaks = 0:2 / 2, limits = 0:1) +
  scale_color_viridis_d(option = "A") +
  scale_fill_viridis_d(option = "A") +
  facet_wrap(~ actor, labeller = label_both, nrow = 1)
```

This is yet another reminder there was very little variation across blocks.

### Posterior prediction for new clusters.

Here's how we might make Figure 13.7, left. Note how we've dropped the `a[actor]` and `g[block]` terms from `spread_draws()`, and added in `a_bar`. For this plot, we didn't need `d_pred` either.

```{r}
#| fig-width: 2
#| fig-height: 3

p1 <- m13.4 |> 
  spread_draws(a_bar, b[treatment]) |> 
  mutate(p = plogis(a_bar + b)) |> 
  
  ggplot(aes(x = treatment, y = p)) +
  stat_lineribbon(point_interval = mean_qi, .width = 0.80, 
                  fill = "gray67") +
  scale_x_continuous(NULL, labels = treatment_labels_vec) +
  scale_y_continuous("proportion pulled left", limits = 0:1) +
  facet_wrap(~ "average actor")

p1
```

If we want to depict the variability across the chimps, we need to bring `sigma_a` into the calculations. In the block of code, below, we simulate a bundle of new intercepts defined by

$$\text{simulated chimpanzees} \sim \operatorname{Normal}(\bar \alpha, \sigma_\alpha).$$

If we are going to continue using a `spread_draws()`-based approach, to my mind it makes sense to first convert the data to a wider format with respect to `treamtment`. Then we sample from `rnorm()`, and then we can convert the data back to the long format with respect to `treatment`. Otherwise you'd have different draws for different levels of `.draw`, which I believe would bungle the results.

```{r}
#| fig-width: 2
#| fig-height: 3

set.seed(13)

p2 <- m13.4 |> 
  spread_draws(a_bar, sigma_a, b[treatment]) |> 
  pivot_wider(names_from = treatment, values_from = b) |> 
  mutate(a_sim = rnorm(n = n(), mean = a_bar, sd = sigma_a)) |> 
  pivot_longer(cols = `1`:`4`, values_to = "b") |> 
  mutate(treatment = as.integer(name)) |> 
  mutate(p = plogis(a_sim + b)) |> 
  
  ggplot(aes(x = treatment, y = p)) +
  stat_lineribbon(point_interval = mean_qi, .width = 0.80, 
                  fill = "gray67") +
  scale_x_continuous(labels = treatment_labels_vec) +
  scale_y_continuous(NULL, breaks = NULL, limits = 0:1) +
  facet_wrap(~ "marginal of actor")

p2
```

To simulate actors for the left panel of Figure 13.7, our basic data-wrangling workflow will be similar to what we just did. This time we'll use the `ndraws` argument within `spread_draws()` to randomly sample a subset of the 4,000 post-warmup posterior draws in the posterior, rather than use them all. When we plot, instead of summarizing the results with `stat_lineribbon()`, we just plot individual lines with `geom_line()`, and we use the `.draw` index as a stand-in for simulated actors.

```{r}
#| fig-width: 2
#| fig-height: 3

set.seed(13)

p3 <- m13.4 |> 
  spread_draws(a_bar, sigma_a, b[treatment], ndraws = 100) |> 
  pivot_wider(names_from = treatment, values_from = b) |> 
  mutate(a_sim = rnorm(n = n(), mean = a_bar, sd = sigma_a)) |> 
  pivot_longer(cols = `1`:`4`, values_to = "b") |> 
  mutate(treatment = as.integer(name)) |> 
  mutate(p = plogis(a_sim + b))  |> 
  
  ggplot(aes(x = treatment, y = p, group = .draw)) +
  geom_line(alpha = 1/2, linewidth = 1/4) +
  scale_x_continuous(NULL, labels = treatment_labels_vec) +
  scale_y_continuous(NULL, breaks = NULL, limits = 0:1) +
  facet_wrap(~ "100 simulated actors")

p3
```

Here's the complete version of Figure 13.7.

```{r}
#| fig-width: 6
#| fig-height: 3

p1 | p2 | p3
```

### Post-stratification.

## ~~Summary~~ Centered, non-centered, and the secret third parameterization {#sec-Centered-non-centered-and-the-secret-third-parameterization}

I'd like to spend some more time talking about how we parameterize multilevel models. To that end, let's load the `reedfrogs` data once more.

```{r}
data(reedfrogs, package = "rethinking")
d <- reedfrogs |>
  # Add a the `tank` cluster variable
  mutate(tank = 1:n())
rm(reedfrogs)

# What?
glimpse(d)
```

Again we make the `stan_data` version of the data for **rstan**.

```{r}
stan_data <- d |>
  select(surv, density, tank) |> 
  compose_data()

# What?
str(stan_data)
```

In @sec-Example-Multilevel-tadpoles when we fit the multilevel reedfrogs model, we used the centered parameterization. The equation for that model was

$$
\begin{align*}
\text{surv}_i & \sim \operatorname{Binomial}(n_i, p_i) \\
\operatorname{logit}(p_i) & = \alpha_i \\
\alpha_i    & \sim \operatorname{Normal}(\bar \alpha, \sigma) \\
\bar \alpha & \sim \operatorname{Normal}(0, 1.5) \\
\sigma      & \sim \operatorname{Exponential}(1),
\end{align*}
$$

where this time for simplicity sake I'm just calling what McElreath denoted $\alpha_{\text{tank}[i]}$ as $\alpha_i$, and I'm also replacing McElreath's somewhat unnecessarily confusing prior notation $\alpha_j$ with $\alpha_i$.^[I should acknowledge there are many other cases where McElreath's more elaborate notation would be necessary. But in this special case, it's not and I'm concerned it would get in the way of the basic points I'm trying to make in this section.] We didn't fit a non-centered version of this model, but if we had, I believe we could have described it in statistical notation as

$$
\begin{align*}
\text{surv}_i & \sim \operatorname{Binomial}(n_i, p_i) \\
\operatorname{logit}(p_i) & = \alpha_i \\
\alpha_i & = \color{blue}{\bar \alpha + z_i \sigma} \\
\bar \alpha       & \sim \operatorname{Normal}(0, 1.5) \\
\color{blue}{z_i} & \sim \color{blue}{\operatorname{Normal}(0, 1)} \\
\sigma            & \sim \operatorname{Exponential}(1),
\end{align*}
$$

for which we have taken cues from the statistical notation McElreath used for `m13.4nc` on page 424 in the text, and I then repeated in @sec-Non-centered-chimpanzees, above. We'll fit this version of the model with **rstan** in a bit.

There are other ways to parameterize the model, though. For a nice exposition, Gelman and Hill discussed at least five in Section 12.5 their [-@gelmanDataAnalysisUsing2006] text. The one I'd like to introduce comes most directly from the notation used throughout @singerAppliedLongitudinalData2003, which is the text from which I first learned about multilevel models.^[You can also find this notation in Equation 12.6 in @gelmanDataAnalysisUsing2006.] I don't know what to call this parameterization, but to stick with the language McElreath has used in this chapter, we'll tentatively dub it the *semi-centered* parameterization,^[Go [here](https://x.com/SolomonKurz/status/1833922891588796644) for a nice Twitter discussion on this parameterization. Doctoral candidate [Francisco Garre-Frutos](https://franfrutos.github.io/) was the first person to suggest the term. We even got the official Stan account to weigh in on the topic (e.g., [here](https://x.com/mcmc_stan/status/1833941708112789872)).] which is

$$
\begin{align*}
\text{surv}_i & \sim \operatorname{Binomial}(n_i, p_i) \\
\operatorname{logit}(p_i) & = \alpha_i \\
\alpha_i & = \bar \alpha + \color{blue}{u_i} \\
\color{blue}{u_i} & \sim \color{blue}{\operatorname{Normal}(0, \sigma)} \\
\bar \alpha       & \sim \operatorname{Normal}(0, 1.5) \\
\sigma            & \sim \operatorname{Exponential}(1),
\end{align*}
$$

where the new $\color{blue}{u_i}$ terms are deviations around the grand mean $\bar \alpha$, but unlike the $z_i$ deviation terms in the previous model, the $\color{blue}{u_i}$ terms are not given a unit-normal prior. Rather, they are distributed as Gaussian with a fixed mean at zero, but a standard deviation $\color{blue} \sigma$, which is itself estimated from the data. In the language of Singer and Willett, you might think of the first two lines of the equation at the level-1 likelihood, and the third and fourth lines as the level-2 likelihood. The rest are priors.

Since we're going to be comparing these three parameterizations, I'm going to first sample more draws from the original centered `m13.2` to help reduce the effect of MCMC sampling variation.

```{r}
#| message: false
#| results: "hide"

m13.2 <- sampling(
    object = m13.2@stanmodel,
    data = stan_data,
    iter = 8500, warmup = 1000,
    cores = 4, seed = 13)
```

Here we make the `model_code_13.2nc` and `model_code_13.2sc` strings for the non-centered and semi-centered version of the model.

```{r}
model_code_13.2nc <- '
data {
  int<lower=1> n;                       
  array[n] int tank;
  array[n] int density;
  array[n] int surv;
}
parameters {
  real abar;
  vector[n] z;
  real<lower=0> sigma;
}
model {
  surv ~ binomial(density, inv_logit(abar + z[tank] * sigma));
  
  abar ~ normal(0, 1.5);
  z ~ normal(0, 1);
  sigma ~ exponential(1);
}
generated quantities {
  vector[n] log_lik;
  for (i in 1:n) log_lik[i] = binomial_lpmf(surv[i] | density[i], inv_logit(abar + z[tank[i]] * sigma));
  
  vector[n] a;
  a = abar + z * sigma;
}
'

model_code_13.2sc <- '
data {
  int<lower=1> n;                       
  array[n] int tank;
  array[n] int density;
  array[n] int surv;
}
parameters {
  real abar;
  vector[n] u;
  real<lower=0> sigma;
}
model {
  surv ~ binomial(density, inv_logit(abar + u));  // Level-1 likelihood
  u ~ normal(0, sigma);                           // Level-2 likelihood
  abar ~ normal(0, 1.5);
  sigma ~ exponential(1);
}
generated quantities {
  vector[n] a;
  a = abar + u;
  
  vector[n] log_lik;
  for (i in 1:n) log_lik[i] = binomial_lpmf(surv[i] | density[i], inv_logit(a[i]));
}
'
```

Now sample from the non-centered and semi-centered posteriors with `stan()`.

```{r}
#| echo: false

# save(m13.2nc, file = "fits/m13.2nc.rda")
# save(m13.2sc, file = "fits/m13.2sc.rda")
load(file = "fits/m13.2nc.rda")
load(file = "fits/m13.2sc.rda")
```
    
```{r}
#| eval: false

m13.2nc <- stan(
  model_code = model_code_13.2nc, 
  data = stan_data,
  iter = 8500, warmup = 1000,
  cores = 4, seed = 13)

m13.2sc <- stan(
  model_code = model_code_13.2sc, 
  data = stan_data,
  iter = 8500, warmup = 1000,
  cores = 4, seed = 13)
```

For the sake of space, I'm not going to show the `print()` output for these. But if you're following along on your computer, it's work giving a look.

```{r}
#| eval: false

print(m13.2, pars = c("abar", "sigma", "a[1]", "a[48]"), probs = c(0.055, 0.945))
print(m13.2nc, pars = c("abar", "sigma", "a[1]", "a[48]"), probs = c(0.055, 0.945))
print(m13.2sc, pars = c("abar", "sigma", "a[1]", "a[48]"), probs = c(0.055, 0.945))
```

Here instead we'll compare the three parameterizations by the major model parameters, $\bar a$, $\sigma$ and $a_i$, with a coefficient plot.

```{r}
#| fig-width: 6
#| fig-height: 7
#| warning: false

par_vec <- c("abar", "sigma", str_c("a[", 1:9, "]"), str_c("a[", 10:48, "]"))

bind_rows(
  as_draws_df(m13.2),
  as_draws_df(m13.2nc),
  as_draws_df(m13.2sc)
) |> 
  select(par_vec) |> 
  mutate(parameterization = rep(c("centered", "non-centered", "semi-centered"), each = n() / 3)) |> 
  pivot_longer(-parameterization, names_to = "par") |> 
  mutate(par = factor(par, levels = par_vec)) |> 
  
  ggplot(aes(x = value, y = par, color = parameterization)) +
  stat_pointinterval(.width = 0.89, 
                     linewidth = 1/4, point_size = 1/2, position = position_dodge(-0.7)) +
  scale_color_viridis_d(option = "C", end = 0.7) +
  labs(x = "posterior",
       y = NULL)
```

They're all the same with MCMC simulation error.

Here we compare the three versions of the model by their HMC convergence statistics in a faceted dot plot.

```{r}
#| fig-width: 8
#| fig-height: 3

bind_rows(
  summarise_draws(m13.2, default_convergence_measures()),
  summarise_draws(m13.2nc, default_convergence_measures()),
  summarise_draws(m13.2sc, default_convergence_measures())
) |> 
  filter(variable %in% par_vec) |> 
  mutate(parameterization = rep(c("centered", "non-centered", "semi-centered"), each = n() / 3)) |> 
  pivot_longer(rhat:ess_tail, names_to = "stat") |> 
  
  ggplot(aes(x = value, y = fct_rev(parameterization), 
             color = parameterization, fill = parameterization)) +
  stat_dots() +
  scale_y_discrete(NULL, expand = expansion(mult = 0.1)) +
  scale_color_viridis_d(option = "C", end = 0.7, breaks = NULL) +
  scale_fill_viridis_d(option = "C", end = 0.7, breaks = NULL) +
  facet_wrap(~ stat, scales = "free_x")
```

I'm not interested in making claims about one being better than the others, but it's good to have options, and I'm glad to add one more to the mix. I've been low-key chewing on this for like five years, and it feels good to finally clarify my thoughts in words and code.

```{r}
#| echo: false
#| eval: false
#| warning: false

# They're almost identical by WAIC
loo_compare(
  extract_log_lik(m13.2) |> waic(),
  extract_log_lik(m13.2nc) |> waic(),
  extract_log_lik(m13.2sc) |> waic()
) |> 
  print(simplify = FALSE)
```

## Session info {-}

```{r}
sessionInfo()
```

## Comments {-}

