# Preface {.unnumbered}

## What and why {-}

This book is based on the second edition of Richard McElreath's [-@mcelreathStatisticalRethinkingBayesian2020] text, [*Statistical rethinking: A Bayesian course with examples in R and Stan*](https://xcelab.net/rm/statistical-rethinking/). My contributions show how to fit the models he covered with [**rstan**](https://mc-stan.org/rstan/) [@R-rstan], which allows one to fit Bayesian models in **R** [@R-base] using the Hamiltonian Monte Carlo algorithms provided by Stan [@standevelopmentteamStanUserGuide2024]. I also prefer plotting and data wrangling with the packages from the [**tidyverse**](https://www.tidyverse.org/) [@R-tidyverse; @wickhamWelcomeTidyverse2019], so we'll be using those methods, too.

This is a sibling book of my translation of McElreath's second edition into **brms** code [@kurzStatisticalRethinkingSecondEd2023], which you can find by clicking [here](https://bookdown.org/content/4857/).

## How this compares with the **brms** translation, and what this means for readers {-}

I think **rstan** is harder to use than **brms**. Its syntax is more technical, and it doesn't follow many of the conventions used by other popular statistics packages, such as **stats** and **lme4**. **rstan** is also more general than **brms**, allowing for a variety of unconventional and bespoke models. Therefore I'm presuming the readership for this book will be intermediate to advanced in both statistical skills, and general **R** programming skills. I will still take a pedagogical tone in this book, but my pace will be a little faster than in my other books.

Also, because this is sibling book to my **brms** translation, I will not replicate all my previous work. Rather, the code in this book will be more tightly focused on reproducing the models and primary analyses with an **rstan** + **tidyverse** framework. The supporting prose will be sparser, and I will have fewer quotes from the text. If you want a richer engagement with McElreath's text, you probably want my **brms** translation instead.

## **R** setup {-}

To get the full benefit from this ebook, you'll need some software. Happily, everything will be free (provided you have access to a decent personal computer and an good internet connection).

First, you'll need to install **R**, which you can learn about at [https://cran.r-project.org/](https://cran.r-project.org/).

Though not necessary, your **R** experience might be more enjoyable if done through the free RStudio interface, which you can learn about at [https://posit.co/products/open-source/rstudio/](https://posit.co/products/open-source/rstudio/).

Once you have installed **R**, execute the following to install the bulk of the add-on packages. This may take a few minutes to finish. Go make yourself a coffee.

```{r}
#| eval: false

packages <- c("bayesplot", "dagitty", "devtools", "GGally", "ggdag", "patchwork", "posterior", "remotes", "rstan", "tidybayes", "tidyverse", "tigris")

install.packages(packages, dependencies = TRUE)
```

A couple of the other packages are not officially available via the Comprehensive R Archive Network (CRAN; [https://cran.r-project.org/](https://cran.r-project.org/)). You can download them directly from GitHub by executing the following.

```{r}
#| eval: false

devtools::install_github("stan-dev/cmdstanr")
devtools::install_github("rmcelreath/rethinking")
```

It's possible you'll have problems installing some of these packages. Here are some likely suspects and where you can find help:

* for difficulties installing **cmdstanr**, go to [https://mc-stan.org/cmdstanr/articles/cmdstanr.html](https://mc-stan.org/cmdstanr/articles/cmdstanr.html);
* for difficulties installing **rethinking**, go to [https://github.com/rmcelreath/rethinking#installation](https://github.com/rmcelreath/rethinking#installation); and
* for difficulties installing **rstan**, go to [https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started).

## Conventions  {-}

As in my other books, I use a handful of formatting conventions gleaned from [*R4DS*](https://r4ds.had.co.nz/introduction.html#running-r-code), [*The tidyverse style guide*](https://style.tidyverse.org/) [@wickhamTidyverseStyleGuide2020], and [*R markdown: The definitive guide*](https://bookdown.org/yihui/rmarkdown/software-info.html) [@xieMarkdownDefinitiveGuide2020].

* **R** code blocks and their output appear in a gray background. E.g.,

```{r}
2 + 2 == 5
```

* **R** and the names of specific package (e.g., **rstan**) are in **boldface** font.
* Functions are in a typewriter font and followed by parentheses, all atop a gray background (e.g., `stan()`).
* When I want to make explicit the package a given function comes from, I insert the double-colon operator `::` between the package name and the function (e.g., `posterior::as_draws_df()`).
* **R** objects, such as data or function arguments, are in typewriter font atop gray backgrounds (e.g., `chimpanzees`, `.width = 0.89`).
* You can detect hyperlinks by their typical [blue-colored font](https://solomonkurz.netlify.app/).

For the **rstan** code, I'm also taking cues from the [*Stan Program Style Guide*](https://mc-stan.org/docs/stan-users-guide/style-guide.html) section of the *Stan User’s Guide*. However, as I'm relatively new to **rstan**, don't be surprised if you find some quirks an inconsistencies.

## Warning  {-}

This book is a side-product of another project for which I needed to learn more about **rstan**. It was also a good opportunity to learn more about making a [Quarto](https://quarto.org/)-based ebook [@quarto2024]. For this 0.0.3 release, I'm only up through Chapter 13. I will continue to chip away on this as my schedule permits, but I am setting no goals and no deadlines. On the one hand, it would be nice to have another finished book in my portfolio; on the other hand, I really don't like using **rstan** and I still much prefer a **brms**-based workflow. The remaining chapters in this book may find themselves fleshed out in quick succession, or this whole project may lay fallow. We'll see...

## Comments! {-}

This is my first ebook with comments sections! Please play nice. I don't plan on heavily monitoring the comments, but if needed I reserve the right to delete, block, and so on. 

## Thank-you's are in order {-}

I'd like to thank the following for their helpful contributions:

* Vincent Arel-Bundock ([\@vincentarelbundock](https://github.com/vincentarelbundock)),
* Francisco Garre-Frutos([\@franfrutos](https://github.com/franfrutos)),
* Andrew Heiss ([\@andrewheiss](https://github.com/andrewheiss)),
* Andrew Johnson ([\@andrjohns](https://github.com/andrjohns)),
* John K. Kruschke ([\@kruschke](https://github.com/kruschke)), and
* Jeff Pollock ([\@jeffpollock9](https://github.com/jeffpollock9)).

Many of these folks have shared **rstan** code that a directly helped make this book possible. I especially appreciate and recommend Arel-Bundock's website [*Statistical Rethinking 2 with Stan and R*](https://vincentarelbundock.github.io/rethinking2/), wherein you can find rendered `Rmarkdown` files with loads of **rstan** code on many of the same problems I'm tackling in this book. If you compare our works, you'll see I have been heavily influenced by Arel-Bundock's **rstan** style.

## License and citation {-}

This book is licensed under the Creative Commons Zero v1.0 Universal license, the details for which you can about [here](https://github.com/ASKurz/Statistical_Rethinking_2_with_rstan/blob/main/LICENSE). In short, you can use my work. Just make sure you give me the appropriate credit the same way you would for any other scholarly resource. Here's the basic citation information:

```{r}
#| eval: false

@book{kurzStatisticalRethinking2rstan2024,
  title = {Statistical rethinking 2 with rstan and the tidyverse},
  author = {Kurz, A. Solomon},
  year = {2024},
  month = {september},
  edition = {version 0.0.3},
  url = {https://solomon.quarto.pub/sr2rstan/},
  doi = {https://doi.org/10.5281/zenodo.13755933}
}
```

